---
title: "Predicting plant population trend from photosynthetic rate"
subtitle: "EDS222"
author: "Sam Muir"
date: "December 12, 2023"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
    css: ['default', 'metropolis', 'metropolis-fonts', 'my-css.css']
---

```{r setup, include=FALSE}
library(tidyverse)
library(BIEN)
library(rredlist)
library(janitor)
library(here)
library(sjPlot)
library(generics)
library(patchwork)
library(kableExtra)

## leaf photosynthetic rate per leaf area data ##
lpr <- BIEN_trait_trait("leaf photosynthetic rate per leaf area") %>%
  rename(scientific_name = scrubbed_species_binomial)  %>%
  mutate(trait_value = as.numeric(trait_value)) %>%
  filter(trait_value < 70)
# there's a few values there in the 200,000+ which don't make sense physiologically...
#they're also all from the same study, so there may be some incorrect calculations in the data.
#These values were removed for simplicity

## find avg photosynthetic rate for each species ##
lpr_summary <- lpr %>%
  group_by(scientific_name) %>%
  summarise(avg_photo = mean(trait_value, na.rm = TRUE))

## read in redlist data ##
redlist <- read_csv(here("data", "redlist_species_data/assessments.csv")) %>%
  clean_names() %>%
  filter(population_trend %in% c("Stable", "Increasing", "Decreasing")) %>% # remove data deficient observations
  # create new pop trend groupings
  mutate(pop_trend = case_when(population_trend == "Decreasing" ~ "Decreasing",
                               population_trend %in% c("Stable", "Increasing") ~ "Stable_Inc"),
         # binary values for pop trend
         pop_trend_binary = case_when(pop_trend == "Decreasing" ~ 1,
                                      pop_trend == "Stable_Inc" ~ 0))


#setdiff(lpr$scrubbed_species_binomial, redlist$scientific_name)

## join lpr with the redlist data, keeping species present in both dfs ##
join <- inner_join(lpr_summary, redlist)

## threat data ##
# climate threatened speices
climate_threatened <- read_csv(here("data", "IUCN_climate_threatened.csv")) %>%
  mutate(climate_threatened = "yes")

# human threatened species
# this includes both "human intrusions & disturbance" and "residential & commercial development" catagories from IUCN
human_res <- read_csv(here("data","IUCN_human_res.csv")) %>%
  mutate(human_threatened = "yes")

## threat joins ##
# those not climate threatened will be catagorized as "no" under the climate_threatened col
climate_join <- left_join(join, climate_threatened) %>%
  mutate(climate_threatened = replace(climate_threatened, is.na(climate_threatened), "no"))

# those not human threatened will be catagorized as "no" under the climate_threatened col
clim_human_res <- left_join(climate_join, human_res) %>%
  mutate(human_threatened = replace(human_threatened, is.na(human_threatened), "no"))
```

# Background

### Plants are threatened by many factors**

- Climate change
- Habit loss
- Human disturbance
- Natural Disasters

--
### Factors that make certain species more resiliant
- Traits (high range tolerance)
- Associations with symbiotic fungi
- Enhanced/ efficient metabolic processes

--
### Endangered and Threatened Species

- number of plants listed as endangered
- number with decreasing population as listed by IUCN


---
# Question:

## Can we predict is a plant species will have a decreasing based on its photosynthetic rate and if it is threatened by climate change or human activity?

---

# Analysis Plan

- Logistic Regression

##### $$\operatorname{logit}(p)=\log \left(\frac{p}{1-p}\right)=\beta_0+\beta_1  (Photosynthetic Rate) + \beta_2  (Climate) + \beta_3  (Human) +\varepsilon $$

- Odds Ratio

##### $$\operatorname{odds}(\hat{p})=\frac{\hat{p}}{1-\hat{p}}=\exp \left(\hat{\beta}_0+\hat{\beta}_1 \cdot x\right)$$

- Make Predictions ?

```{r}
threat_prob <- function(b0, b1, b2, b3, photo, climate, human) {
  equation <- b0 + b1 * photo + b2 * climate + b3 * human
  prob <- (exp(equation)) / (1 + exp(equation))
  print(prob)
}
```

---
# Data

### Photosynthetic Rate
- `BIEN` package in R
  - Botanical Information and Ecology Network (BIEN) was developed by collaborating ecologists at the National Center for Ecological Analysis and Synthesis (NCEAS)
- Population Trend and Threats
  - IUCN Red List
  
---
# First looks

```{r echo=FALSE, fig.height=6}
ggplot(data = join, aes(x = avg_photo, y = pop_trend)) + 
  geom_jitter(width = 0, height = 0.05, alpha = 0.8) +
  theme_linedraw() +
  labs(x = "Avg. Photosynthetic Rate per Leaf Area (Âµmol m-2 s-1)", 
       y = "Population Trend")
```

---
# Logistic Regression
```{r echo=FALSE, fig.height=6}
# binary resp glm plot
ggplot(join, aes(x = avg_photo, y = pop_trend_binary)) + 
  geom_jitter(width = 0, height = 0.05, alpha = 0.8) +
  geom_smooth(method = "glm", se = FALSE, color = "red", 
              method.args = list(family = "binomial")) +
  theme_linedraw() +
  labs(x = "Avg. Photosynthetic Rate per Leaf Area (Âµmol m-2 s-1)", 
       y = "Has Decreasing Population Trend")
```
---
# Logistic Regression
```{r echo=FALSE}
# model
mod_photo <- glm(pop_trend_binary ~ avg_photo, 
                  data = join, 
                  family = "binomial") # binomial dist; prob of 1 and 0 over many trials

# model table
sjPlot::tab_model(mod_photo,
          transform = NULL,
          pred.labels = c("Intercept", "Avg. Photosynthetic Rate (Âµmol m-2 s-1)"), # predictor labels
          show.p = TRUE, # include p-val
          p.style = c("numeric_stars"),
          p.threshold = c(0.10, 0.05, 0.01),
          dv.labels = c("log Probability of Decreasing Population"),
          string.p = "p-value",
          show.r2 = FALSE,
          title = "Logisitc Regression Model Results for Photosynthetic Rate",
          digits = 3)
```

$$\operatorname{logit}(\hat p)=\log \left(\frac{\hat p}{1-\hat p}\right)=-0.464-0.089 x$$

---
# Odds Ratio

By how much does the model predict that the odds of having a decreasing population will change with each additional unit in photosynthetic rate?
--
$$\operatorname{odds}(\hat{p})=\frac{\hat{p}}{1-\hat{p}}=\exp \left(\hat{\beta}_0+\hat{\beta}_1 \cdot x\right)$$
--
```{r}
exp(mod_photo$coefficients[2])
```
--
The model estimates that one additional unit in photosynthetic rate is associated with a change in the odds ratio of $e^{-0.089} = 0.915$, or a 8.9% decrease in the odds of having a decreasing population.

---
# Odds Ratio
```{r echo=FALSE}
# compute odds hat var
photo_fitted <- mod_photo %>%
  augment(type.predict = "response") %>%
  mutate(y_hat = .fitted) %>%
  mutate(odds_hat = y_hat / (1 - y_hat)) # y_hat is p

# plot
ggplot(photo_fitted, aes(x = avg_photo, y = odds_hat)) +
  geom_point(size = 1) + 
  geom_line() + 
  scale_y_continuous("Log-odds of being threatened") +
  labs(x = "Avg. Photosynthetic Rate (Âµmol m-2 s-1)") +
  theme_linedraw()
```
---
# Logistic Model

<small>$$\operatorname{logit}(p)=\log \left(\frac{p}{1-p}\right)=\beta_0+\beta_1  (Photosynthetic Rate) + \beta_2  (Climate) + \beta_3  (Human) +\varepsilon$$<small>

--

```{r}
mod <- glm(pop_trend_binary ~ avg_photo + climate_threatened + human_threatened,
           data = clim_human_res,
           family = "binomial")

exp(coef(mod))

mod$coefficients[2]
```

---
# Logistic Model

```{r echo=FALSE}
tab_model(mod,
          transform = NULL,
          pred.labels = c("Intercept", "Average Photosyntheitc Rate (Âµmol m-2 s-1)", 
                          "Climate Threatened", "Human Threatened"),
          p.style = c("numeric_stars"),
          p.threshold = c(0.10, 0.05, 0.01),
          show.p = TRUE,
          string.p = "p-value",
          dv.labels = c("log Threat Pobability"),
          show.r2 = FALSE,
          title = "Logisitc Regression Model Results for Photosynthetic Rate, Climate Threatened, and Human Threatened",
          digits = 3)
```

--
Based on this model:

- Plants threatened by climate factors have their odds of having a decreasing population increase by a factor of 4.29
- Plants threatened by human factors have their odds increase by a factor of 2.52
- The odds of having a decreasing population trend fall by about 10% for each unit increase in photosynthetic rate.

---
# Making Probalistic Predictions

```{r}
threat_prob <- function(b0, b1, b2, b3, photo, climate, human) {
  equation <- b0 + b1 * photo + b2 * climate + b3 * human
  prob <- (exp(equation)) / (1 + exp(equation))
  print(prob)}
```

```{r echo=FALSE, results='hide', warning=FALSE, message=FALSE}
# coefficients
b0 <- mod$coefficients[1] #Intercept
b1 <- mod$coefficients[2] #photo
b2 <- mod$coefficients[3] #climate
b3 <- mod$coefficients[4] #human

# Function for testing probabilities
threat_prob <- function(b0, b1, b2, b3, photo, climate, human) {
  equation <- b0 + b1 * photo + b2 * climate + b3 * human
  prob <- (exp(equation)) / (1 + exp(equation))
  print(prob)
}

# find plants not on red list with the lowest photosynthetic rates
lpr_pred_low <- anti_join(lpr_summary, redlist) %>%
  arrange(desc(avg_photo)) %>%
  tail(5)

lpr_pred <- anti_join(lpr_summary, redlist) %>%
  arrange(desc(avg_photo)) %>%
  head(5) %>%
  rbind(lpr_pred_low)

# binary climate and human values come from literature search
pred_cols <- data.frame(climate_threatened = c(0,0,1,1,0,1,0,1,1,0),
                        human_threatened = c(0,0,0,0,1,0,0,1,0,0)) 

lpr_pred <- cbind(lpr_pred, pred_cols)

for (i in seq_along(lpr_pred$scientific_name)) {
  lpr_pred$y_hat[i] <- threat_prob(b0, b1, b2, b3, 
                                       lpr_pred$avg_photo[i],
                                       lpr_pred$climate_threatened[i],
                                       lpr_pred$human_threatened[i])
}
```

```{r echo=FALSE}
lpr_pred %>% 
  arrange(desc(avg_photo)) %>%
  kbl(linesep = "", booktabs = TRUE, col.names = c("Speices", "Photosynthetic Rate", "Climate", "Human", "ˆy")) %>%
  kable_styling(bootstrap_options = c("striped", "condensed"),
                latex_options = c("striped", "hold_position"), full_width = TRUE)
```





